Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-04-12 paper/paper.tex
Restructure and more corrections

\begin{abstract}
We propose a novel Bayesian approach for functional linear or logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subset of the RKHS generated by the underlying covariance function. By imposing a suitable prior distribution on such a RKHS, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated from Markov Chain Monte Carlo (MCMC) methods. Several prediction methods derived from this posterior distribution turn out to be competitive against other usual  alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.
\end{abstract}

@@

\begin{comment}
Posibles temas de desarrollo teórico:
  \begin{itemize}
    \item Enfoque de procesos Gaussianos como priors en modelos RKHS, eligiendo un kernel \(K'\) que tenga buenas propiedades con respecto a \(K\). Distribuciones conjugadas.
    \item Temas de consistencia de la posteriori: concentración hacia el ``verdadero'' valor el parámetro o bien convergencia hacia la ``verdadera'' posteriori.
    \item Distancia máxima entre las funciones de \(\Hcal_{0,p}(K)\) y las de \(\Hcal_0(K)\) (ó \(\Hcal(K)\)).
    \item Robustez de las distribuciones ante pequeñas perturbaciones.
    \item  En \citet[Sec.~2.5]{grollemund2019bayesian}, los autores proponen un estimador Bayesiano para sus funciones ``escalón'' \(\beta_\theta(\cdot)\) utilizando una pérdida cuadrática, de la forma
  \[
    \hat \beta \in \arg\min_{d\in L^2[0, 1]} \int\int (\beta_\theta(t) - d(t))^2 \pi(\theta\mid \mathcal D_n)\, dt\, d\theta.
  \]
  ¿Podríamos hacer algo parecido con nuestro parámetro funcional \(\alpha \in \Hcal_{0,p}(K)\)?
  \item Considerar oras distribuciones a priori (especialmente en regresión logística).
  \end{itemize}
\end{comment}

@@

The problem of predicting a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle we could simply regard the functional data as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on continuous data. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} and \citet{horváth2012inference}.

FDA is undoubtedly an active area of research; accordingly, there are many recent contributions on how to tackle functional data problems, both from a theoretical and practical standpoint. Chief among them is the approach of reducing the problem to a finite-dimensional one, for example using a truncated basis expansion or spline interpolation methods. At the same time, much effort has also been put into the opposite task: generalizing different concepts to the infinite-dimensional framework. Examples of this endeavor include a functional Mahalanobis distance \citep{berrendero2020mahalanobis} or a functional partial least squares algorithm \citep{delaigle2012methodology}. Another technique found in the related literature is the use of Gaussian processes to model the functional behavior of the data \citep[see for instance][]{shi2011gaussian}. These ideas extend the theory of Gaussian processes in classical finite-dimensional regression settings \citep[e.g.][]{rasmussen2004gaussian}, providing an alternative non-parametric Bayesian approach to inference and prediction problems.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some more subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will further suppose without loss of generality that \(X\) is centered, that is, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

@@

 The most common functional linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the book by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is an error term independent from \(X\) with \(\E [\epsilon]=0\), and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). In this case, the inference on \(\beta\) is hampered by the fact that \(L^2[0,1]\) is an extremely wide space that contains many non-smooth or ill-behaved functions, so that any estimation procedure involving optimization on it would typically be very hard. In spite of this, model~\eqref{eq:l2-linear-model} is not flexible enough to include ``simple'' finite-dimensional models based linear combinations of the marginals, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta(\cdot)\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2021functional}.

It turns out that in both circumstances a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it includes a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is particularly appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018use, berrendero2021functional} for the case of logistic regression. Incidentally, these models also shed light on the near-perfect classification phenomenon for functional data, described by \citet{delaigle2012achieving} and further examined in the works of \citet{berrendero2018use} or \citet{torrecilla2020optimal}.

A major aim of this work is to motivate these recently-proposed models inside the functional framework, while also providing efficient techniques to apply them in practice. Our main contribution is the proposal of a Bayesian approach to parameter estimation within the aforementioned RKHS models, in which a prior distribution is imposed on the unknown functional parameter to obtain a posterior distribution after seeing the data. In this way, one can incorporate pre-existing beliefs about the parameters, thus increasing the flexibility of the estimation procedure. Although setting a prior distribution on a function space is generally a hard task, the specific parametric formulation of the RKHS models we propose greatly facilitates this (see Section~\ref{sec:methodology} for details). A similar Bayesian approach has been recently explored in \citet{grollemund2019bayesian}, albeit not within a RKHS framework. \incomment{Comentar resultados experimentales}

Another set of techniques extensively studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_j)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we could easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_j\}\), and thus provide a Bayesian-motivated variable selection process along with the other prediction methods that naturally arise within our model. In this way, in addition to making predictions about the input data, we could evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also appeared recently in the literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration. \incomment{Comentar resultados experimentales}

@@

As we can see, the covariance function of \(X\) (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS. An integral operator closely related to this kernel is the so-called covariance operator, namely \(\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds\) for \(f \in L^2[0, 1]\), which is self-adjoint and compact when \(K\) is continuous \citep[e.g.][Th.~4.6.2]{hsing2015theoretical}. It is worth mentioning that this operator provides several alternative definitions of \(\Hcal(K)\); for example, the RKHS can be identified with the image of the square root of the covariance operator, i.e., \(\Hcal(K) = \mathcal K^{1/2}(L^2[0, 1])\), with inner product \(\dotprod{f}{g}_K = \dotprod{\mathcal K^{-1/2}(f)}{\mathcal K^{-1/2}(g)}\). Furthermore, we can think of the norm in \(\Hcal(K)\) as an \(L^2\)-like regularized norm, since it is easy to show that we can also express \(\Hcal(K) = \{f \in L^2[0, 1]: \ \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}^2 < \infty \}\), where \(\lambda_j\) and \(\phi_j\) are the eigenvalues and eigenfunctions of \(\mathcal K\), respectively. In this case, the corresponding inner product is \(\dotprod{f}{g}_K = \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}\dotprod{g}{\phi_j}\). Note that, since the spectral theorem for compact operators tells us that the sequence of eigenvalues \(\{\lambda_j\}\subset \R^+_0\) tends to zero \citep[e.g.][Th.~4.2.4]{hsing2015theoretical}, this definition highlights the fact that functions in \(\Hcal(K)\) are smooth, in the sense that their components in an orthonormal basis need to vanish quickly.

@@

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration are those obtained by taking a functional parameter \(\alpha \in \Hcal(K)\) and replacing the scalar product for \(\dotprod{x}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}. For practical and computational reasons, the value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_1,\dots,\beta_p \in \R, \ t_1,\dots,t_p \in [0, 1]\right\}.
\end{equation}

The general idea will be to impose a prior distribution on these free parameters (and possibly others) to eventually derive a posterior model after incorporating the available sample information.  Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(\Psi\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition, we can write \(\dotprod{x}{\alpha}_K \equiv \sum_j \beta_j x(t_j)\).

In view of~\eqref{eq:h0p}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\beta_j\) and the times \(t_j\) separately. Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\Hcal(K)\) merely by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\). Even though we actually work on \(\Hcal_{0,p}(K)\), the parameter \(p\) can be selected in several meaningful ways that still make use of the available data, and the range of possibilities is not very wide in practice.

@@

It is worth mentioning that the model remains linear in the sense that it fundamentally involves a random variable belonging to the linear span of the process \(X\) in \(L^2(\Omega)\). Also, note that given the time instants \(t_j\) the model becomes a multiple linear model with the \(X(t_j)\) as scalar covariates. As a matter of fact, this RKHS model is particularly suited as a basis for variable selection methods, and also entails the classical \(L^2\)-model~\eqref{eq:l2-linear-model} under certain conditions \citep[see][Sec.~3]{berrendero2020general}. In addition, as far as generalizations go the proposed model is easily extensible, seeing as how several functional covariates could be added to obtain an extended model such as \(Y=\alpha_0 + \Psi_{X^{1}}(\alpha_1) + \cdots + \Psi_{X^{q}}(\alpha_q) + \epsilon\). In that case, as argued in \citet{grollemund2019bayesian} for a similar situation, if we were to set a prior distribution on all the parameters involved, we could recover the full posterior by looking alternately at the posterior distribution of each covariate conditional on the rest of them.

@@

As for the posterior distribution, we compute a function proportional to its log-density, since that is all that is needed for the MCMC algorithm to work. A standard algebraic manipulation in~\eqref{eq:bayes-theorem} yields the following result.

@@

Lastly, in order to generate predictions let us recall that when performing the empirical posterior approximation, on each of the \(M\) steps of the iterative MCMC algorithm we get an approximate sample \(\theta^{(m)*}=(b^{(m)*}, \tau^{(m)*}, \alpha_0^{(m)*}, (\sigma^2)^{(m)*})\) of the posterior distribution \(\pi(\theta| \mathcal D_n)\). Assuming now a previously unseen test set \(\mathcal D'_{n'}\) in the same conditions as \(\mathcal D_n\), we propose to construct three different kinds of predictors based on the MCMC samples, each of them following a different strategy.

@@

In much the same way as the linear regression model described before, this RKHS-based logistic regression model partially solves some of the issues with the \(L^2\)-model~\eqref{eq:l2-logistic-model} stated in Section~\ref{sec:intro}. Moreover, it negates the problem of non-existence of the MLE in the functional case, since our proposed model is finite-dimensional\footnote{Even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist. However, this problem can be easily circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.}. Additionally, under specific conditions this model also entails the logistic \(L^2\)-model, and in fact it does hold under mild assumptions whenever the conditional distributions \(X | Y=i\) (\(i=0,1\)) are homoscedastic Gaussian processes \citep[see Theorem 1 in][]{berrendero2021functional}.

@@

\bibliographystyle{ba}
\bibliography{bibliography}

>>>>

\begin{abstract}
We propose a novel Bayesian parametric approach for functional linear and logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated from Markov Chain Monte Carlo (MCMC) methods. Several prediction strategies derived from this posterior distribution turn out to be competitive against other usual alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.
\end{abstract}

@@

\begin{comment}
Posibles temas de desarrollo teórico:
  \begin{itemize}
    \item Enfoque de procesos Gaussianos como priors en modelos RKHS, eligiendo un kernel \(K'\) que tenga buenas propiedades con respecto a \(K\). Distribuciones conjugadas.
    \item Temas de consistencia de la posteriori: concentración hacia el ``verdadero'' valor el parámetro o bien convergencia hacia la ``verdadera'' posteriori.
    \item Distancia máxima entre las funciones de \(\Hcal_{0,p}(K)\) y las de \(\Hcal_0(K)\) (ó \(\Hcal(K)\)).
    \item Robustez de las distribuciones ante pequeñas perturbaciones.
    \item  En \citet[Sec.~2.5]{grollemund2019bayesian}, los autores proponen un estimador Bayesiano para sus funciones ``escalón'' \(\beta_\theta(\cdot)\) utilizando una pérdida cuadrática, de la forma
  \[
    \hat \beta \in \arg\min_{d\in L^2[0, 1]} \int\int (\beta_\theta(t) - d(t))^2 \pi(\theta\mid \mathcal D_n)\, dt\, d\theta.
  \]
  ¿Podríamos hacer algo parecido con nuestro parámetro funcional \(\alpha \in \Hcal_{0,p}(K)\)?
  \item Considerar otras distribuciones a priori (especialmente en regresión logística).
  \end{itemize}
\end{comment}

@@

The problem of predicting a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle the functional data could be simply regarded as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on data varying over a continuum. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} or the book by \citet{horvath2012inference}.

FDA is undoubtedly an active area of research, which finds applications in a wide variety of fields, such as biomedicine, finance, meteorology or chemistry \citep[see][]{ullah2013applications}. Accordingly, there are many recent contributions on how to tackle functional data problems, both from a theoretical and practical standpoint. Chief among them is the approach of reducing the problem to a finite-dimensional one, for example using a truncated basis expansion or spline interpolation methods \citep{aguilera2013comparative}. At the same time, much effort has also been put into the opposite task: generalizing different concepts to the infinite-dimensional framework. Examples of this endeavor include a functional Mahalanobis distance \citep{berrendero2020mahalanobis} or a functional partial least squares algorithm \citep{delaigle2012methodology}. Another technique found in the related literature is the use of Gaussian processes to model the functional behavior of the data \citep[see for instance][]{shi2011gaussian}. These ideas extend the theory of Gaussian process regression in classical finite-dimensional settings \citep[e.g.][]{rasmussen2004gaussian}, providing an alternative non-parametric Bayesian approach to inference and prediction problems. Additional non-parametric methods for functional prediction and classification were notably explored in \citet{ferraty2006nonparametric}.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some more subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

@@

 The most common functional linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\), and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). In this case, the inference on \(\beta\) is hampered by the fact that \(L^2[0,1]\) is an extremely wide space that also contains many non-smooth or ill-behaved functions, so that any estimation procedure involving optimization on it would typically be very hard. In spite of this, model~\eqref{eq:l2-linear-model} is not flexible enough to include ``simple'' finite-dimensional models based on linear combinations of the marginals, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2021functional}.

It turns out that in both circumstances a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it includes a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is especially appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018use, berrendero2021functional} for the case of logistic regression. Incidentally, these models also shed light on the near-perfect classification phenomenon for functional data, described by \citet{delaigle2012achieving} and further examined in the works of \citet{berrendero2018use} and \citet{torrecilla2020optimal}.

A major aim of this work is to motivate these recently-proposed models inside the functional framework, while also providing efficient techniques to apply them in practice. Our main contribution is the proposal of a Bayesian approach to parameter estimation within the aforementioned RKHS models, in which a prior distribution is imposed on the unknown functional parameter to obtain a posterior distribution after seeing the data. Although setting a prior distribution on a function space is generally a hard task, the specific parametric formulation of the RKHS models we propose greatly facilitates this (see Section~\ref{sec:methodology} for details). A similar Bayesian approach has been recently explored in \citet{grollemund2019bayesian}, albeit not within a RKHS framework. \incomment{Comentar nuestros resultados experimentales}

Another set of techniques extensively studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_j)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we could easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_j\}\), and thus provide a Bayesian-motivated variable selection process along with the other prediction methods that naturally arise within our model. In this way, in addition to making predictions about the input data, we could evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also appeared recently in the literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration. \incomment{Comentar nuestros resultados experimentales}

@@

As we can see, the covariance function of \(X\) (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS. An integral operator closely related to this kernel is the so-called covariance operator, namely \(\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds\) for \(f \in L^2[0, 1]\), which is self-adjoint and compact when \(K\) is continuous \citep[e.g.][Th.~4.6.2]{hsing2015theoretical}. It is worth mentioning that this operator provides several alternative definitions of \(\Hcal(K)\); for example, the RKHS can be identified with the image of the square root of the covariance operator, i.e., \(\Hcal(K) = \mathcal K^{1/2}(L^2[0, 1])\), with inner product \(\dotprod{f}{g}_K = \dotprod{\mathcal K^{-1/2}(f)}{\mathcal K^{-1/2}(g)}\). Furthermore, we can think of the norm in \(\Hcal(K)\) as an \(L^2\)-like regularized norm, since it is easy to show that we can also express \(\Hcal(K) = \{f \in L^2[0, 1]: \ \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}^2 < \infty \}\), where \(\lambda_j\) and \(\phi_j\) are the eigenvalues and (orthonormal) eigenfunctions of \(\mathcal K\), respectively. In this case, the corresponding inner product is \(\dotprod{f}{g}_K = \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}\dotprod{g}{\phi_j}\). Note that, since the spectral theorem for compact operators tells us that the sequence of eigenvalues \(\{\lambda_j\}\) tends to zero \citep[e.g.][Th.~4.2.4]{hsing2015theoretical}, this definition highlights the fact that functions in \(\Hcal(K)\) are smooth, in the sense that their components in an orthonormal basis need to vanish quickly.

@@

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration are those obtained by taking a functional parameter \(\alpha \in \Hcal(K)\) and replacing the scalar product for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}. For practical and computational reasons, the value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}

The general idea will be to impose a prior distribution on these free parameters (and possibly others) to eventually derive a posterior model after incorporating the available sample information.  Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition, we can write \(\dotprod{x}{\alpha}_K \equiv \sum_j \beta_j x(t_j)\).

In view of~\eqref{eq:h0p}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\beta_j\) and the times \(t_j\) separately. Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\Hcal(K)\) merely by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\). Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of possible values is not very large in practice.

@@

It is worth mentioning that the model remains linear in the sense that it fundamentally involves a random variable belonging to the linear span of the process \(X\) in \(L^2(\Omega)\). Also, note that given the time instants \(t_j\), the model becomes a multiple linear model with the \(X(t_j)\) as scalar covariates. As a matter of fact, this RKHS model is particularly suited as a basis for variable selection methods, and also entails the classical \(L^2\)-model~\eqref{eq:l2-linear-model} under certain conditions \citep[see][Sec.~3]{berrendero2020general}. In addition, as far as generalizations go the proposed model is easily extensible, seeing as how several functional covariates could be added to obtain an extended model such as \(Y=\alpha_0 + \Psi_{X^{1}}(\alpha_1) + \cdots + \Psi_{X^{q}}(\alpha_q) + \epsilon\). In that case, as argued in \citet{grollemund2019bayesian} for a similar situation, if we were to set a prior distribution on all the parameters involved, we could recover the full posterior by looking alternately at the posterior distribution of each covariate conditional on the rest of them.

@@

As for the posterior distribution, we only compute a function proportional to its log-density, since that is all that is needed for the MCMC algorithm to work. A standard algebraic manipulation in~\eqref{eq:bayes-theorem} yields the following result.

@@

In order to generate predictions, let us recall that when performing the empirical posterior approximation, on each of the \(M\) steps of the iterative MCMC algorithm we get an approximate sample \(\theta^{(m)*}=(b^{(m)*}, \tau^{(m)*}, \alpha_0^{(m)*}, (\sigma^2)^{(m)*})\) of the posterior distribution \(\pi(\theta| \mathcal D_n)\). Assuming now a previously unseen test set \(\mathcal D'_{n'}\) in the same conditions as \(\mathcal D_n\), we propose to construct three different kinds of predictors based on the MCMC samples, each of them following a different strategy.

@@

In much the same way as the linear regression model described before, this RKHS-based logistic regression model partially solves some of the issues with the \(L^2\)-model~\eqref{eq:l2-logistic-model} stated in Section~\ref{sec:intro}. Moreover, it negates the problem of non-existence of the MLE in the functional case, since our proposed model is finite-dimensional\footnote{Even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist \citep{albert1984existence}. However, this problem can be easily circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.}. Additionally, under specific conditions this model also entails the logistic \(L^2\)-model, and in fact it does hold under mild assumptions whenever the conditional distributions \(X | Y=i\) (\(i=0,1\)) are homoscedastic Gaussian processes \citep[see Theorem 1 in][]{berrendero2021functional}.

@@

\bibliographystyle{bib/ba}
\bibliography{bib/bibliography}
